Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-19 paper-1/manuscript.tex
Undo notation spacing tweaks
Consistent newline use in fig legends
Editing

% Common notation
\renewcommand{\R}[3][]{{}^{#1}_{}\!\mathbf{R}^{#2}_{#3}}
\renewcommand{\T}[3][]{{}^{#1}_{}\mathbf{T}^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\!\mathbf{D}^{#2}_{#3}}
\newcommand{\mean}[1]{\overline{#1}}

@@

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{Kubota2021-dx,Harms2014-ah,Jacobsen2001-sc}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the \dev{} sequence to A in the \std{} sequence). In other words, we used the \msc{} sequence as a neutral baseline where the target stimulus was presented no more often than in the \dev{} sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes fired per neuron in response to target stimulus A in the \dev{} and \msc{} sequences, defining a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{\mean{\R[dev]{A}{}} - \mean{\R[msc]{A}{}}}{\mean{\R[dev]{A}{}} + \mean{\R[msc]{A}{}}} \label{eqn-ddi}
\end{equation}

@@

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two sequences, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in Figure \ref{fig4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both sequences. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, indicating higher \dev{} activity. Since the stimulus was identical in both sequences, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower initial $\theta$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in \dev{} sequence ($\R[dev]{A}{i}$) and \msc{} sequence ($\R[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \R{A}{i}$).
\textbf{b} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

To assess the difference in synaptic currents in a way that allowed direct comparison to $\theta$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\theta$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We quantified the contrast in $\theta$ at the start of target trials as $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$, where $\T{}{i} = \theta_i(t=0)$, corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, and correlated this with the target trial response size contrast $\Delta \R{A}{i}$, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{i}$ tended to show a greater increase in their response to target stimuli.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $\theta$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $\theta$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $\theta$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $\theta$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

Having established that the increased response to A in \dev{} trials was driven by lower thresholds before stimulation, we next turned to the cause of this reduction. A priori, since threshold adaptation is a direct reflection of activity, and since the response to target trials was larger in \dev{} than in \msc{}, the lower thresholds in \dev{} must have been caused by lower activity in non-target trials, i.e. lower responses to B in \dev{} than to B through E in \msc{}. Yet, since activity is modulated by thresholds as shown above, and since the dev sequence was random, implying that $\T[dev]{A}{} \approx \T[dev]{B}{}$, we might expect the lowered thresholds in \dev{} to yield not only higher target responses, but also higher non-target responses. To resolve this apparent paradox, we will turn to an analysis of the spatial arrangement of activity and adaptation.

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\T[dev]{}{}$ is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{A}{i}$, i.e., $\theta_i$ at the start of A trials (mean over trials), in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

First, we mapped the average value of $\theta$ at the start of target trials, $\T{A}{i}$, to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the \msc{} sequence, $\T[msc]{A}{i}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression across all excitatory neurons. Neurons are colored by latency rank as in panel a. Additionally, excitatory neurons among the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{d} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{e} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we found, for each neuron, the minimum response time (i.e., time to first spike) across B trials in all sequences, and ranked neurons by this response time. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast, $\Delta \R{N}{i}$, to the contrast in $\theta$ at the start of B trials, $\Delta \T{B}{i}$. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated $\Delta \R{N}{i}$ to $\Delta \T{A}{i}$. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red, left plot) increasing their response in the \dev{} sequence, while the response of the later portion of the network (blue, right plot) decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$ were robustly positively correlated (median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast ($\Delta \R{N}{}$, Figure \ref{fig6}c) and the median contrast in $\T{B}{}$ ($\Delta \T{B}{}$, Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} (T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). $\Delta \T{B}{}$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).

Finally, to show that high $\T[dev]{B}{early}$ was causal in reducing the response to B, we correlated the median contrast in $\T{}{}$ among the early portion, $\Delta \T{B}{early}$, with the network-wide median contrast of the response to B, $\Delta \R{B}{}$, as shown in Figure \ref{fig6}e. Across networks and stimuli, we found a strong negative correlation ($\rho$ = -0.587, p = 9.56e-13, Wald test, n = 120), indicating that greater TA in the neurons responding early was indeed predictive of a lowered response to B in the overall network.

@@

Thus, we have finally traced the greater response to A in the \dev{} sequence to its roots, and revealed two complementary roles of TA: The frequent presentation of non-target stimulus B in the \dev{} sequence and concomitant increased activity in a subset of neurons responding early to B causes \emph{local suppression} by TA in this subset, leading to smaller network responses. By contrast, stimuli in the \msc{} sequence are presented more sporadically, allowing local suppression to decay, thus leading to larger network responses. This difference in non-target response sizes -- small in \dev{}, large in \msc{} -- is then reflected in network-wide \emph{global suppression}, such that the network is more excitable to stimulation with A (or other infrequent inputs) in the \dev{} sequence than in the control context.

@@

We had good reasons for this assumption: A priori, updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B may have been able to lead to a build-up of depression strong enough to suppress its own response.

To show in the sample network that the model does not maintain a history beyond the most recent trial, we correlated the number of spikes in \dev{} target trials with $x_j$ at the start of later trials across excitatory neurons. We found that, while the correlation was strong and very clearly positive in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; Wald test, n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial. Similar behavior was also seen in the \msc{} sequence, as shown in Figure \ref{sfig-STD-correlation-decay}.

To show that STD being carried over between repeated trials was responsible for the reduction of non-target \dev{} responses relative to \msc{}, however, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) was affected by changes in STD. Unlike $\T{}{j}$, which directly affects the neuron to which the value belongs, $x_j$ exerts its effects on the other side of a synapse. Therefore, we devised a measure $\D{}{}$ to estimate the postsynaptic impact of STD in terms of membrane potential lost, mirroring the effect of $\T{}{}$ in the sense that decreasing $V_m$ and increasing the firing threshold both increase the depolarization necessary to make a neuron fire.

>>>>

% Common notation
\renewcommand{\R}[3][]{{}^{#1}_{}\mathbf{R}^{#2}_{#3}}
\renewcommand{\T}[3][]{{}^{#1}_{}\mathbf{T}^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\mathbf{D}^{#2}_{#3}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\Rhat}[3][]{{}^{#1}_{}\mathbf{\widehat{R}}^{#2}_{#3}}

@@

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{Kubota2021-dx,Harms2014-ah,Jacobsen2001-sc}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the \dev{} sequence to A in the \std{} sequence). In other words, we used the \msc{} sequence as a neutral baseline where the target stimulus was presented no more often than in the \dev{} sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes fired per neuron in response to target stimulus A in the \dev{} and \msc{} sequences, denoted as $\mean{\R[dev]{A}{}}$ and $\mean{\R[msc]{A}{}}$, respectively, and define a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{\mean{\R[dev]{A}{}} - \mean{\R[msc]{A}{}}}{\mean{\R[dev]{A}{}} + \mean{\R[msc]{A}{}}} \label{eqn-ddi}
\end{equation}

@@

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two sequences, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in Figure \ref{fig4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both sequences. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, which indicate higher \dev{} activity. Since the stimulus was identical in both sequences, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower initial $\theta$.
\textbf{a} Post-stimulus spike histogram across target trials in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in the \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $\theta$ at the start of target trials ($\Delta \T{A}{i}$) and the contrast in the average number of spikes fired in target trials ($\Delta \R{A}{i}$), and associated linear regression.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

To assess the difference in synaptic currents in a way that allowed direct comparison to $\theta$, we refer to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\theta$ alone to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We quantified the contrast in $\theta$ at the start of target trials as $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$, where $\T{}{i} = \theta_i(t=0)$, corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, and correlated this with the target trial response size contrast $\Delta \R{A}{i}$ across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} ($\Delta \T{A}{exc} < 0$: T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{i}$ tended to show a greater increase in their response to target stimuli.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $\theta$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $\theta$ and $V_m$ by the magnitude of the response difference and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $\theta$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $\theta$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

Having established that the increased response to A in \dev{} trials was driven by lower thresholds before stimulation, we next turned to the cause of this reduction. A priori, since threshold adaptation is a direct reflection of activity, and since the response to target trials was larger in \dev{} than in \msc{}, the lower thresholds in \dev{} must have been caused by lower activity in non-target trials, i.e., by lower responses to B in \dev{} than to B through E in \msc{}. Yet, since activity is modulated by thresholds as shown above, and since the dev sequence was random, implying that $\T[dev]{A}{} \approx \T[dev]{B}{}$, we might expect the lowered thresholds in \dev{} to yield not only higher target responses, but also higher non-target responses. To resolve this apparent paradox, we will turn to an analysis of the spatial arrangement of activity and adaptation.

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower thresholds in \dev{} are a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

First, we mapped the average value of $\theta$ at the start of target trials, $\T{A}{i}$, to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the \msc{} sequence, $\T[msc]{A}{i}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, a small number of neurons near the non-target stimulation site (B) were strongly adapted, whereas most of the network was less adapted, consistent with the data shown above.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression. Neurons are colored by latency rank as in panel a. Additionally, excitatory neurons among the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{d} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{e} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we ranked each neuron by the minimum response time (i.e., time to first spike) across B trials in all sequences. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast, $\Delta \R{N}{i}$, to the contrast in $\theta$ at the start of B trials, $\Delta \T{B}{i}$. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated $\Delta \R{N}{i}$ to $\Delta \T{A}{i}$. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red, left plot) increasing their response in the \dev{} sequence, while the response of the later portion of the network (blue, right plot) decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$ were robustly positively correlated (median Pearson's $\rho$ 0.992: T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast ($\Delta \R{N}{}$, Figure \ref{fig6}c) and the median contrast in $\T{B}{}$ ($\Delta \T{B}{}$, Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} ($\Delta \R{B}{early} > 0$: T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased ($\Delta \R{B}{late} < 0$: T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). $\Delta \T{B}{}$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).

Finally, to show that high $\T[dev]{B}{early}$ was causal in reducing the response to B, we correlated the median contrast in $\T{B}{}$ among the early portion, $\Delta \T{B}{early}$, with the network-wide median contrast of the response to B, $\Delta \R{B}{}$, as shown in Figure \ref{fig6}e. Across networks and stimuli, we found a strong negative correlation ($\rho$ = -0.587, p = 9.56e-13, Wald test, n = 120), indicating that greater TA in the neurons responding early was indeed predictive of a lowered response to B in the overall network.

@@

Thus, we have finally traced the greater response to A in the \dev{} sequence to its roots and revealed two complementary roles of TA: The frequent presentation of non-target stimulus B in the \dev{} sequence and concomitant increased activity in a subset of neurons responding early to B causes \emph{local suppression} by TA in this subset, leading to smaller network responses. By contrast, stimuli in the \msc{} sequence are presented more sporadically, allowing local suppression to decay, thus leading to larger network responses. This difference in non-target response sizes -- small in \dev{}, large in \msc{} -- is then reflected in network-wide \emph{global suppression}, such that the network is more excitable to stimulation with A (or other infrequent inputs) in the \dev{} sequence than in the control context.

@@

We had good reasons for this assumption: Updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B may have been able to lead to a build-up of depression strong enough to suppress its own response.

To show in the sample network that the model does not maintain a history beyond the most recent trial, we correlated $\R[dev]{A}{j}$ with $x_j$ at the start of later trials across excitatory neurons. We found that, while the correlation was strong and very clearly positive in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; Wald test, n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial. Similar behavior was also seen in the \msc{} sequence, as shown in Figure \ref{sfig-STD-correlation-decay}.

To show that STD being carried over between repeated trials was responsible for the reduction of non-target \dev{} responses relative to \msc{}, however, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) was affected by changes in STD. Unlike $\theta_j$, which directly affects the activity of neuron $j$, $x_j$ only exerts its effects postsynaptically. Therefore, we devised a measure $\D{}{}$ to estimate the postsynaptic impact of STD in terms of membrane potential lost, mirroring $\T{}{}$ in the sense that decreasing $V_m$ and increasing the firing threshold both increase the depolarization necessary to make a neuron fire.

A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. In order to respect this notion while at the same time minimizing the influence of activity differences between sequences, we used the average response $\Rhat{S}{j} = \frac{1}{3} (\R[dev]{S}{j} + \R[msc]{S}{j} + \R[std]{S}{j})$ of a neuron $j$ to a given stimulus $S$ as a measure of the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at $V_m = V_{rest}$ (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $\Rhat{S}{j} x_j^S k$, where we ignored added depression due to repeated spikes within a trial and instead defined $x_j^S$ as the mean depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in each neuron $i$ as
\begin{equation}
    \D{S}{i} = \sum_{j \in pre} \Rhat{S}{j} (1-x_j^S) k \label{eqn-D}
\end{equation}
Notice that $\D{}{}$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $\T{}{}$, albeit more approximate.
